---
title: "Immerse Interview Prep"
author: "Fredrick Kakembo"
date: "2022-05-7"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loading the data

```{r}
library(ggplot2)
library(dplyr)

data <- read.csv("applicant_dataset_v2.csv")

```

# Check the size of the data

```{r}
dim(data)

colnames(data)
```


```{r}
#Deal with the categorical data to factors
data$agecat <- as.factor(data$agecat)
data$siteid <- as.factor(data$siteid)
data$BSdich <- as.factor(data$BSdich)
data$LAMP <- as.factor(data$LAMP)
data$anymalaria <- as.factor(data$anymalaria)
```

```{r}
attach(data)
```


```{r}
head(data)
```

```{r}
# id column

#data$id
length(data$id)
length(unique(data$id))
```

### 1. Age Category

```{r}
table(agecat)

sum(is.na(agecat))

p <- ggplot(data = data, aes(agecat)) +
  geom_bar(fill="steelblue") + ylim(c(0,600))

#Reordering the categories to start with 6months
p + scale_x_discrete(limits = c("6 months - < 5 years","5 years - < 11 years" ,">= 18 years")) +
  geom_text(stat='count', aes(label=..count..), vjust=-1)+
  theme_bw()
```

### 2. Site ID

```{r}
#Site
table(siteid)
sum(is.na(siteid))  #None missing

ggplot(data = data, aes(siteid)) + geom_bar(fill="steelblue") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5)+
  theme_bw()+ ylim(c(0,400))
```


### 3. blood smear result
- 1 is positive, 0 is negative

```{r}
table(BSdich) # 1 is positive and 0 is negative

sum(is.na(BSdich)) #Number of missing

ggplot(data = data, aes(BSdich)) + geom_bar(fill="steelblue") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5) +
  theme_bw() + ylim(c(0,815))
```


### 4. LAMP
loop-mediated isothermal amplification result for submicroscopic P falciparum parasitemia. LAMP was only performed if blood smear was negative. 1 is positive, 0 is negative

```{r}
library(dplyr)
#LAMP
data_0 <- subset(data, BSdich == 0) #Subset data for only when BS is neg 
dim(data_0)
#View(data_0)

table(data_0$LAMP)

sum(is.na(data_0$LAMP)) #

ggplot(data = data_0, aes(LAMP)) + geom_bar(fill="steelblue") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5)+
  theme_bw() + ylim(c(0,400))


### Just to be sure, LAMP was only performed when BSdich was negative
data_1 <- subset(data, BSdich == 1) 
table(data_1$LAMP)
```


### 5. anymalaria
diagnosis of malaria was made on that date (clinical symptoms + positive blood smear). “1” is a diagnosis (case) of malaria, “0” is no malaria 

```{r}
table(anymalaria)
sum(is.na(anymalaria))

ggplot(data = data, aes(anymalaria)) + geom_bar(fill="steelblue") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5)+
  theme_bw() + ylim(c(0,1010))
```

## 1a. Site and age
Do we have an even balance of the age categories sampled from each region?

```{r}
#SiteID and Agecat
test2 <-   data %>%
  group_by(siteid, agecat) %>%
  summarize(t2.len = length(agecat)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = siteid, y = t2.len, fill = agecat)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") + ylim(0,220)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "SiteID", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd", "#2ca25f")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

## 2a Age category and BSdich
Are people of a given age category likely to be positive for a BS

```{r}
# Agecat and BSdich
test2 <-   data %>%
  group_by(agecat, BSdich) %>%
  summarize(t2.len = length(BSdich)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = agecat, y = t2.len, fill = BSdich)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") + ylim(0,400)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Age Category", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

Confirm if there exists a difference; 

H0: There is no link / relationship between age category and results of blood smear.

H1: There is a link between age category and results of blood smear.

```{r}
table(agecat, BSdich)
c = chisq.test(agecat,BSdich)
c
c$expected
c$p.value
```

Since p-value is significant, we have enough evidence to reject the null hypothesis and conclude that there is a link (relationship) between age category and results of blood smear.

~~ P-value is significant. We are likely to have most positive cases among 4-11 years, followed by more cases in 6months-5 years and least in individuals above 18 years ~~

 Keep in mind that "statistically significant" does not always imply "meaningful" when using the chi-square test.



##  2b agecat and anymalaria


```{r}
# Agecat and anymalaria
test2 <-   data %>%
  group_by(agecat, anymalaria) %>%
  summarize(t2.len = length(anymalaria)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = agecat, y = t2.len, fill = anymalaria)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") + ylim(0,510)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Age Category", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

Test significance if there is any relationship between Agecat and anymalaria

```{r}
table(agecat, anymalaria)
c = chisq.test(agecat,anymalaria)
c
c$expected
c$p.value
```

P-value is still significant, hence we can conclude there is a relationship between age category and one having malaria. 


## 3 a  siteid and BS

```{r}
# siteid and BSdich
test2 <-   data %>%
  group_by(siteid, BSdich) %>%
  summarize(t2.len = length(BSdich)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = siteid, y = t2.len, fill = BSdich)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") +  ylim(0,300)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Site Location", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

# Check if this difference is significant

```{r}
table(siteid, BSdich)
c = chisq.test(siteid,BSdich)
c
c$expected
c$p.value
```

P-value is significant meaning there is a relationship between site location of a person and the result of the blood smear. Its more likely that people from Kanungu and Tororo will have positive blood smear than people from Jinja. 

What about malaria diagnosis?

## 3 b. siteid and anymalaria


```{r}
# siteid and anymalaria
test2 <-   data %>%
  group_by(siteid, anymalaria) %>%
  summarize(t2.len = length(anymalaria)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = siteid, y = t2.len, fill = anymalaria)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") + ylim(0,350)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Site Location", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

Confirm if the observed difference is significant

```{r}
table(siteid, anymalaria)
c = chisq.test(siteid,anymalaria)
#c
c$expected
c$p.value
```

p-value is significant, hence we conclude that there is a relation between site site location and someone having malaria. 


## ageCat vs LAMP

Qn: Are individuals of a given age category likely be have submicroscopic results

```{r}
detach(data)
# Agecat and LAMP
test2 <-   data_0 %>%
  group_by(agecat, LAMP) %>%
  summarize(t2.len = length(LAMP)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = agecat, y = t2.len, fill = LAMP)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") +  ylim(0,250)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Age Category", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```

Alot of data is missing from the group >=18

Test the significance if any

```{r}
table(data_0$agecat, data_0$LAMP)
c = chisq.test(data_0$agecat, data_0$LAMP)
#c
c$expected
c$p.value
```

Its significance but level of significance is not that too strong like the previous




## siteID vs LAMP


```{r}
# siteid and LAMP
test2 <-   data_0 %>%
  group_by(siteid, LAMP) %>%
  summarize(t2.len = length(LAMP)) %>%
  mutate(t2.prop = round(t2.len / sum(t2.len) * 100, 1))

p22 <- ggplot(test2, aes(x = siteid, y = t2.len, fill = LAMP)) +  
  geom_bar( stat = "identity", position = position_dodge(width = 0.9) , color="black") +  ylim(0,160)  +
  geom_text(aes(label = paste(t2.prop, "%", sep = "")), 
            position = position_dodge(width = 0.9), vjust = -0.8) + theme_bw() + 
  theme(legend.position = "bottom") + labs(x = "Site Location", y = "Frequency Count") +
  scale_fill_manual(values = c("#deebf7", "#3182bd")) +
  theme( axis.text=element_text(size=11),text=element_text(size=12))

p22
```


```{r}
table(data_0$siteid, data_0$LAMP)
c = chisq.test(data_0$siteid,data_0$LAMP)
c
c$expected
c$p.value
```

P-value is still significant


```{r}
dim(na.omit(data_0))

s <- na.omit(data_0[c("siteid", "LAMP")])
dim(s)
```

```{r}
table(s$siteid, s$LAMP)
c = chisq.test(s$siteid,s$LAMP)
c
c$expected
c$p.value

```


Note a limitatiob of chi square, When Sample size is small, the test is less trustworthy. However, with very large sample sizes, even relatively trivial relationships may be declared statistically significant. 




















